\Ace: An Actively-Typed Compilation Environment\\for High-Performance Computing

### 2012-10-14 ace/ace.tex
ace: Revised through section 2.C

\begin{abstract}
We introduce the Ace compilation environment and demonstrate its suitability as a foundational tool for both research and practice in high-performance computing.
Ace consists of a statically-typed programming language with {\em user-extensible semantics} defined using a compile-time meta\-language, Python. Users specify new primitive types and their operations by equipping type definitions, which are first-class objects in the metalanguage, with  methods that the compiler selectively invokes when type checking and translating expressions, a mechanism we call {\em active type-checking and translation (\ATT)}. We demonstrate the flexibility of this mechanism by implementing primitives from several widely-known languages, including the entirety of OpenCL. Ace also supports more general forms of metaprogramming, and Ace functions can be compiled and launched directly from Python with standard numeric data structures as arguments. Using each of these features, we designed a scientific simulation framework that allows users to modularly specify, compile and orchestrate the execution of parameterized families of scientific simulations on clusters of GPUs. This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing initial evidence that Ace is useful in practice today.
\end{abstract}

@@

Computer-aided simulation and data analysis techniques have transformed science and engineering. Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}. Most of this software targets desktop hardware and about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}. To fully harness the power of these platforms, however, these {\em professional end-user developers} \cite{segal2007some} must write parallel programs.

@@

The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important concerns related to performance, portability, verifiability and ease-of-use. Unfortunately, professional end-users are hesitant to adopt these, with many generally skeptical regarding the practicality of new approaches that originate in the research community. This viewpoint was perhaps most succinctly expressed by a participant in a recent study by Basili et al. \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by researchers working to advance the practice of high-performance computing.

A major barrier to adoption of new language features and abstractions is that they are often developed and distributed together with a new programming language. Some prominent examples of this language-oriented approach include:
\begin{itemize}
\item UPC \cite{upc}, CAF \cite{caf}, Fortress \cite{fortress}, X10 \cite{x10} and Chapel \cite{chapel}, which support different forms of the partitioned global address space (PGAS) model; 
\item Actors, CSPs and other forms of message passing are best supported by languages like Erlang, Charm++, Occam and Axum.
\item Nested data parallelism is best supported by languages like Cilk and NESL.
\end{itemize}

Unfortunately, using a specialized research language comes with a host of practical challenges, such as the cost of learning the language, and the lack of well-developed libraries and tools. In fact, the most widely adopted parallel programming abstractions have been those that are distributed as libraries, such as MapReduce, Global Arrays and MPI. This approach is more practical because it allows developers to progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize different abstractions within a single program without encountering friction inherent to language boundaries. Abstractions developed as libraries are limited, however, in that they cannot alter or augment the semantics of the host language itself. Thus, abstractions that require new type checking rules and compilation logic for the types and associated operations they introduce are at a relative disadvantage in practice.

Ace aims to solve this problem by allowing developers to define and distribute new first-class language constructs in a safely composable manner directly within user libraries using a mechanism called {\em active typechecking and translation (\ATT)}. In particular, \ATT~allows library authors to extend the semantics of the  language while retaining a fixed but flexible grammar. Aided by a form of type inference, the Ace grammar -- a large subset of Python's -- is flexible enough that we are able to define all of the primitives of the C99 and OpenCL languages in a natural manner. These primitives can then be supplemented with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}, as desired. Notably, these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler for a language.

%To enable this and other, Ace programs contain both compile-time and run-time program logic. The compile-time logic, written in the {\em metalanguage}, is the top-level code in each file. Types are first-class objects in the metalanguage, and the logic needed to typecheck and compile the operators associated with a type is bundled with the type definition itself. Abstractions can target existing backend languages, using its existing compilation infrastructure for low-level optimizations, so only the high-level semantics need to be encoded by researchers. Consistent with its design intent, we demonstrate the use of OpenCL as a flexible backend language and compiler framework \cite{opencl}.
%Here, users can define and import new primitive types and operators, as well as do other types of metaprogramming tasks like generate functions programmatically, perform arbitrary code transformations.
% Ace also permits an interactive mode of use whereby the functions can also be executed immediately after being defined, a common usage scenario for professional end-users.

Although the \ATT~mechanism, a key feature of Ace, is designed for use by researchers and domain experts, Ace is designed to be highly practical for professional end-user developers as well. It is based on the Python programming language, which enjoys broad adoption in this community. Indeed, the language itself is implemented entirely as a Python library, so many existing tools, such as editor modes, documentation tools and style checkers, are available immediately.
Ace programs can be generated programmatically and called directly from Python, with numeric arrays as arguments, so Ace integrates immediately with existing workflows to speed up performance-critical portions of code that may benefit from static compilation and support for parallelism.
We demonstrate several of these features by developing a scientific simulation framework, which we use for specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs, in Section \ref{clegans}.

%Researchers design and implement novel parallel programming abstractions\- on a regular basis. Each abstraction aims to make some class of developers more productive by making appropriate trade-offs between performance, portability, verifiability and ease-of-use. It can be observed that the most widely-adopted abstractions to date are those that been implemented as libraries that can be easily deployed within existing, widely-used languages. Abstractions that require introducing new primitive constructs, changing the semantics of existing constructs in particular contexts, or controlling compilation in a fine-grained manner cannot take this approach. This has led to a proliferation of specialized programming languages, each designed around a few privileged abstractions. Unfortunately, such languages have not been widely adopted by practitioners.
%
%slash
%
%Although researchers often propose higher-level language features and parallel programming abstractions that aim to strike a better balance between raw performance, productivity, code portability and verifiability (see Section 2), end-users remain skeptical of new approaches. This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by those in the research community working to advance the practice of scientific and high-performance computing. We begin in Section 2 by developing a set of design and adoption criteria for new languages and abstractions based on prior empirical studies of these {\it professional end-user developers} as well as our observations of characteristics common to successful projects in the past.

%
% Two types of languages see widespread use: productivity-oriented and performance-oriented
% Significant diversity in productivity-oriented languages. Valued for clean syntax and high-level abstractions. Usually very slow. Have a FFI to call into a performance-oriented language.
% Performance-oriented languages -- C or Fortran and their derivatives. Direct memory access, low-level abstractions, more verbose. Portable (though often need to include platform-specific optimizations, which can lead to messy code). Not easily extended with new abstractions and primitives (witness: lots of derivative languages that are not composable because no way to implement these abstractions cleanly)
% Generally don't support verifiability very well (performance-oriented languages have simple type systems); only exception is functional languages.
%

% Simplicity (Errors, Syntax, Core Semantics)
% Extensibility (New Abstractions, Domain-Specific Things, Domain- and Platform-Specific Optimizations)
% Performance (Need to allow control over data layout and movement, perhaps optionally)
% Portability (Architectures and operating systems; support new optimizations)
% Backwards-Compatibility (Existing codebases)
% Verifiability (type systems, metaprogramming-based, formal semantics, small TCB)
% Tool Support (syntax highlighting, editors, debuggers, etc.)
% Social Proof (Will not disappear - open source, significant projects have been developed, community)

%We then introduce a new programming language named Ace (pronounced ``C eloquence'') in Section 3. Ace initially targets code paths where low-level languages, particularly OpenCL, would be used today. By employing a collection of powerful front-end language and compiler design techniques, making pragmatic design choices, and building on top of established infrastructure, we argue that Ace may uniquely satisfy many of the criteria established in Section 2, particularly those related to ease of use, familiarity, performance and tool support.
%
%Unlike many existing languages, Ace is designed to be highly extensible from inception, employing a powerful compile-time language extension and code generation mechanism based around the concept of {\it active libraries} \cite{activelibraries}. We argue that this feature may make Ace particularly suitable as a platform for researchers developing new parallel abstractions and other specialized, high-level language features, as well as  development tools.
%
%In Section 4 we briefly describe a case study where the language was used to conduct large-scale spiking neural circuit simulations on GPU hardware, demonstrating the feasibility of this approach on a non-trivial problem. Finally, conclude in Section 6 with future directions for both the Ace language and the research community.
%

@@

\section{Ace for Professional End-Users}
%Ace is an {\em actively-typed, language-integrated compilation environment}. Let us begin by defining each of these terms. 
%A {\em compilation environment} is a system that allows developers to programmatically control aspects of compilation directly. The language used to do so is called the {\em metalanguage} while the grammar that programs themselves are written in is called the {\em source grammar}. The metalanguage of Ace is Python and the source grammar is a significant subset of Python's own grammar. Programs and metalanguage code can be interleaved within the same file, so Ace is {\em language-integrated}. Ace programs are statically-typed, and types themselves are objects in the metalanguage. They can be equipped by developers with methods that control how operators associated with them are typechecked and compiled. Because of the relationship between this mechanism and the idea of {\em active libraries} \cite{activelibraries}, discussed further in Section \ref{related}, we call them {\em active types}, and thus Ace is {\em actively-typed}.
\subsection{Example 1: Hello, World!}
To demonstrate the structure of programs and libraries in Ace, we begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the OpenCL module distributed with Ace. We note that the OpenCL module, which we use throughout this paper in examples, is simply a library like any other. The core of Ace gives no special treatment to it; it is distributed with Ace for convenience.

Perhaps the most immediately apparent departure from the standard ``Hello, World!'' example comes on lines 3 and 10 of Listing \ref{hello}, which contain \verb|print| statements that are executed at {\em compile-time}. Indeed, Ace programs are Python scripts at the top-level, rather than a list of declarations as in most conventional languages. In other words, Python is the {\em compile-time metalanguage} of Ace. In other words, Ace programs are embedded within Python. By making this choice, Ace can  use Python's well-developed module system and associated distribution infrastructure directly (Line 1).

The \verb|main| function defined on Lines 5-7 contains a single call to the \verb|printf| primitive imported from the \verb|ace.OpenCL| module on Line 1. The function is introduced using the \verb|@OpenCL.fn| decorator, which indicates that it is a {\em actively-typed} Ace function containing run-time logic written for the \verb|OpenCL| backend, also imported on Line 1. Without this decorator, the function would simply be a conventional Python function that could be called only at compile-time (doing so would fail here: \verb|printf| is not a Python primitive.)

On Line 8, the {\em generic function} defined on Lines 5-7, is compiled to produce a {\em concrete function} which we also call \verb|main| (overwriting the previous definition). The distinction between generic and concrete functions will be explained in our next example. For now, we note that only concrete functions are exported when running the Ace compiler, called \verb|acecc|, shown operating at the shell in Listing \ref{helloout}. The \verb|acecc| compiler operates as follows:
\begin{enumerate}
\item Executes the provided Python module (\verb|hello.py|)
\item Produces source code associated with any concrete functions (functions produced using the \verb|compile| method on Line 8 of Listing \ref{hello}) that are in the top-level environment. Each backend may produce one or more files (here, only \verb|hello.cl| is produced).
\end{enumerate}

Use of the \verb|acecc| executable is simply a convenience; the generated code is also available as an attribute of the concrete function: \verb|main.code|. We will also show in Section \ref{pyopencl} that, for backend languages with Python bindings, such as OpenCL and C, functions can be executed directly, without an intermediate compilation step, if desired.

@@

@OpenCL.fn
def map(input, output, f):
    gid = get_global_id(0)
    output[gid] = f(input[gid])
\end{lstlisting}
\caption{[\texttt{listing3.py}] A generic data-parallel higher-order map function written using the user-defined Ace OpenCL module.}
\label{map}
\end{codelisting}

@@

A, B = global_ptr(double), global_ptr(int)
map_add5_dbl = map.compile(A, A, add5.ace_type)
map_add5_int = map.compile(B, B, add5.ace_type)
\end{lstlisting}
\caption{[\texttt{listing4.py}] The generic map function compiled to map the sin function over global arrays of double-precision floating-point numbers. The \texttt{get\_global\_id} and \texttt{sin} functions are OpenCL primitives.}
\label{mapadd5dbl}
\end{codelisting}

@@

__kernel void map_add5_int(
  __global int* input,
  __global int* output)
{
    size_t = gid;
    gid = get_global_id(0);
    output[gid] = add5__1(input[gid]);
}
\end{lstlisting}
\caption{The OpenCL code generated by running \texttt{acecc listing4.py}, or by executing the  \texttt{print listing4.map\_sin\_dbl.code} statement.}
\label{mapout}
\end{codelisting}

The ``Hello, World!'' example demonstrates the basic structure of Ace programs, but it does not require working with types. Listing \ref{map} shows a data-parallel map primitive written using the OpenCL library, introduced above. OpenCL is a data-parallel language -- users can define functions, called {\em kernels}, that execute across thousands of threads. Each kernel has access to its index, called its {\em global id}, which can be used by the programmer to ensure that each thread operates on different parts of the input data. The \verb|map| kernel defined in Listing \ref{map} applies a transfer function, \verb|f|, to the element of the input array, \verb|input| corresponding to its global id, writing the result into the corresponding location in the provided output array, \verb|output|.

As written, \verb|map| is a {\em generic function} (an instance of the class \verb|ace.GenericFn|). This means that it does not have types assigned to its arguments. Indeed, the functionality of \verb|map| can be defined for many combinations of array types for \verb|input| and \verb|output| and functions. In this sense, \verb|map| is actually a family of functions defined for all types assigned to \verb|input|, \verb|output| and \verb|f| for which the operations in the function's body are typesafe. Running \verb|acecc listing3.py| would produce no output, however.

To create a {\em concrete function} (an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler,  types must be assigned to each of the arguments of any externally callable functions. Listing \ref{mapadd5dbl} shows how to use the \verb|compile| method to specialize \verb|map| to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL has a notion of four different memory spaces). Line 9 produces a version specialized for arrays of \verb|double|s and Line 10 produces a version for arrays of \verb|int|s. The output of compilation is shown in Figure \ref{mapout}.

\subsection{Types as Metalanguage Objects}
The types of the arguments of the functions created on Lines 4.9 and 4.10 were written in pure OpenCL would be \verb|__global double*| and \verb|__global int*| respectively. The corresponding types in Ace are \verb|global_ptr(double)| and \verb|global_ptr(int)|, abbreviated as \verb|A| and \verb|B| respectively on Line 8 for convenience (note that here, standard variable assignment in the metalanguage is serving the role that a \verb|typedef| declaration would in a C-like language.)

Types in Ace are values in the metalanguage, Python. More specifically, types are {\em object instances} of user-defined classes that inherit from \verb|ace.Type|. For example, \verb|global_ptr(double)| is an instance of \verb|ace.OpenCL.GlobalPtrType| instantiated with the target type, \verb|double|, as an argument. The type \verb|double| itself (imported from \verb|ace.OpenCL| on Line 4.2) is an instance of \verb|ace.OpenCL.FloatType|. As we will describe further in Section \ref{att}, this notion of ``types as metalanguage objects'' is a key to Ace's flexibility.

>>>>

\begin{abstract}
We introduce the Ace compilation environment and demonstrate its suitability as a foundational tool for both research and practice in high-performance computing.
Ace consists of a statically-typed programming language with {\em user-extensible semantics} specified via a compile-time meta\-language, Python. Users specify new primitive types and their operations by equipping type definitions, which are first-class objects in the metalanguage, with  methods that the compiler selectively invokes when type checking and translating expressions, a mechanism we call {\em active type-checking and translation (\ATT)}. We demonstrate the flexibility of this mechanism by implementing primitives from several widely-known languages, including the entirety of OpenCL, as libraries. Ace also supports more general forms of metaprogramming, and functions can be launched directly from Python with standard numeric data structures as arguments. Using these features, we designed a scientific simulation framework that allows users to modularly specify, compile and orchestrate the execution of parameterized families of scientific simulations on clusters of GPUs. This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing initial evidence that Ace is useful in practice today.
\end{abstract}

@@

Computer-aided simulations and data analysis techniques have transformed science and engineering. Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}. Most of this software targets desktop hardware and about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}. To fully harness the power of these platforms, however, these {\em professional end-user developers} \cite{segal2007some} must write parallel programs.

@@

The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important concerns relating to performance, portability, verifiability and ease-of-use. Unfortunately, professional end-users are hesitant to adopt these, with many being generally skeptical regarding the practicality of new approaches that come from the research community. This viewpoint was perhaps most succinctly expressed by a participant in a recent study by Basili et al. \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by researchers working to advance high-performance computing.

Many of the barriers to adoption of new abstractions stem from the fact that new abstractions are often developed and distributed together with a new programming language. Some prominent examples where this {\em language-oriented approach} has been followed include:
\begin{itemize}
\item UPC \cite{upc}, CAF \cite{caf}, X10 \cite{x10} and Chapel \cite{chapel}, built around first-class support for a partitioned global address space (PGAS) programming model; 
\item Erlang \cite{erlang}, Charm++ \cite{charm}, Occam \cite{occam} and Axum \cite{axum}, built around first-class support for message passing;
\item Cilk \cite{cilk} and NESL \cite{nesl}, built around first-class support for nested data parallelism; and
\item Add something about automatic parallelization of functional programs.
\end{itemize}

There is evidence suggesting that developers would prefer first-class support for a parallel abstraction over a library-based solution, if all else is held constant \cite{ppfirstclass}. Unfortunately, adopting a specialized research language comes with a host of ancillary challenges. Learning a new language can be time-consuming, new languages often lack well-developed libraries and tools, and it can be difficult to port large legacy codebases. Indeed, the most widely adopted parallel programming abstractions have been those that are distributed as libraries for an existing, widely-used language, such as MapReduce, Global Arrays and MPI. This {\em library-oriented approach} has proven more practical because it allows developers to progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize different abstractions within a single program without encountering the friction inherent to language boundaries.

Ace aims to solve this problem by allowing developers to define and distribute new {\em first-class language constructs}, in a safely composable manner, directly within user libraries, via a mechanism we introduce called {\em active typechecking and translation (\ATT)}. In particular, \ATT~is a mechanism that allows developers to  extend the semantics of a language that has a fixed but flexible grammar. The Ace grammar is a large subset of Python's  and we show that it is flexible enough that we are able to define all of the primitives of the C99 and OpenCL languages in a natural manner, aided by a form of type inference. These primitives can then be supplemented with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}. Notably, all of these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler for a language.

Although the \ATT~mechanism, a key feature of Ace, is designed primarily for use by researchers and domain experts, Ace is intended to be highly practical and ready for adoption by professional end-user developers today. It is based on the Python programming language, which enjoys broad adoption in this community. Indeed, the language itself is implemented entirely as a Python library, and many existing tools, such as editor modes, documentation tools and style checkers, can be used without modification.
Ace functions are generated programmatically and can be called directly from Python, with numeric arrays as arguments, so developers can integrate Ace immediately within their existing workflows to define portions of code that may benefit from static compilation and support for parallelism.
We demonstrate the practicality of Ace in Section \ref{clegans} by describing an Ace-based framework that supports specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs.

@@

\section{Ace for Professional End-Users}
%Ace is an {\em actively-typed, language-integrated compilation environment}. Let us begin by defining each of these terms. 
%A {\em compilation environment} is a system that allows developers to programmatically control aspects of compilation directly. The language used to do so is called the {\em metalanguage} while the grammar that programs themselves are written in is called the {\em source grammar}. The metalanguage of Ace is Python and the source grammar is a significant subset of Python's own grammar. Programs and metalanguage code can be interleaved within the same file, so Ace is {\em language-integrated}. Ace programs are statically-typed, and types themselves are objects in the metalanguage. They can be equipped by developers with methods that control how operators associated with them are typechecked and compiled. Because of the relationship between this mechanism and the idea of {\em active libraries} \cite{activelibraries}, discussed further in Section \ref{related}, we call them {\em active types}, and thus Ace is {\em actively-typed}.
\subsection{Example 1: Hello, World!}
To demonstrate the structure of programs and libraries in Ace, let us begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the OpenCL module distributed with Ace. We note that this module, which we use throughout the paper in examples, is simply a library like any other. The core of Ace gives no special treatment to it; it is distributed with Ace for convenience.

Perhaps the most immediately apparent departure from the standard ``Hello, World!'' example comes on lines 3 and 10 of Listing \ref{hello}, which contain \verb|print| statements that are executed at {\em compile-time}. Indeed, Ace programs are Python scripts at the top-level, rather than a list of declarations as in most conventional languages. In other words, Python is the {\em compile-time metalanguage} of Ace; Ace programs are embedded within Python. A consequence of this choice is that Ace can leverage Python's well-developed package system and  distribution infrastructure directly (Line 1).

The \verb|main| function defined on Lines 5-7 contains a single call to the \verb|printf| primitive, imported from the \verb|ace.OpenCL| module on Line 1. The function is introduced using the \verb|@OpenCL.fn| decorator, which indicates that it is a statically-typed Ace function containing run-time logic targeting the \verb|OpenCL| backend, also imported on Line 1. Without this decorator, the function would simply be a conventional Python function that could be called only at compile-time (doing so would fail here: \verb|printf| is not a Python primitive.)

On Line 8, this {\em generic function} defined on Lines 5-7, is compiled to produce a {\em concrete function} that we also identify as \verb|main| (overwriting the previous definition). The distinction between generic and concrete functions will be explained in our next example. For now, we note that only concrete functions are exported when running the Ace compiler, called \verb|acecc|, shown operating at the shell in Listing \ref{helloout}. The \verb|acecc| compiler operates as follows:
\begin{enumerate}
\item Executes the provided Python module (\verb|hello.py|)
\item Produces source code associated with any concrete functions (functions produced using the \verb|compile| method on Line 8 of Listing \ref{hello}) that are in the top-level environment. Each backend may produce one or more files (here, \verb|hello.cl| is produced).
\end{enumerate}

Using the \verb|acecc| executable is simply a convenience; the generated code is also available as an attribute called of the concrete function immediately after it has been defined, accessed as \verb|main.code|. We will show in Section \ref{pyopencl} that, for backend languages with Python bindings, such as OpenCL and C, generic functions can be executed directly, without the intermediate compilation step, if desired.

@@

@OpenCL.fn
def map(input, output, f):
    gid = get_global_id(0)
    output[gid] = f(input[gid])
\end{lstlisting}
\caption{[\texttt{listing3.py}] A generic data-parallel higher-order map function written using OpenCL user module.}
\label{map}
\end{codelisting}

@@

A = global_ptr(double); B = global_ptr(int)
map_add5_dbl = map.compile(A, A, add5.ace_type)
map_add5_int = map.compile(B, B, add5.ace_type)
\end{lstlisting}
\caption{[\texttt{listing4.py}] The generic \texttt{map} function compiled to map the \texttt{add5} function over two  types of input.}
\label{mapadd5dbl}
\end{codelisting}

@@

__kernel void map_add5_int(
  __global int* input,
  __global int* output)
{
    size_t = gid;
    gid = get_global_id(0);
    output[gid] = add5__1(input[gid]);
}
\end{lstlisting}
\caption{[\texttt{listing4.cl}] The OpenCL code generated by running \texttt{acecc listing4.py}.}
\label{mapout}
\end{codelisting}

The ``Hello, World!'' example demonstrates the basic structure of Ace programs, but it does not require working with types. Listing \ref{map} shows an imperative, data-parallel map primitive written using the OpenCL library introduced above. In OpenCL, users can define functions, called {\em kernels}, that execute across thousands of threads. Each kernel has access to a unique index, called its {\em global id}, which can be used by the programmer to ensure that each thread operates on different parts of the input data (Line 5). The \verb|map| kernel defined in Listing \ref{map} applies a transfer function, \verb|f|, to the element of the input array, \verb|input|, corresponding to its global id. It writes the result of this call into the corresponding location in the provided output array, \verb|output|.

As above, \verb|map| is a {\em generic function} (specifically, an instance of the class \verb|ace.GenericFn|). This means that it's arguments have not been assigned types. Indeed, the functionality given by the \verb|map| definition is applicable to many combinations of types for \verb|input| and \verb|output| and functions, \verb|f|. In this sense, \verb|map| is actually a family of functions defined for all types assignments for \verb|input|, \verb|output| and \verb|f| for which the operations in the function's body are well-defined.

Running \verb|acecc listing3.py| would produce no output, however. To create a {\em concrete function} (an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler, types must be assigned to each of the arguments of any externally callable functions. Listing \ref{mapadd5dbl} shows how to use the \verb|compile| method to specialize \verb|map| in two different ways to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL has a notion of four different memory spaces). Line 9 produces a version specialized for arrays of \verb|double|s and Line 10 produces a version for arrays of \verb|int|s. The output of compilation is shown in Figure \ref{mapout}.

\subsection{Types as Metalanguage Objects}
The types of the arguments of the functions being compiled on Lines 4.9 and 4.10, if written directly in OpenCL as in Listing \ref{mapout}, are \verb|__global double*| and \verb|__global int*|, respectively. The corresponding types in Ace are \verb|global_ptr(double)| and \verb|global_ptr(int)|, abbreviated for convenience as \verb|A| and \verb|B| respectively on Line 8 (note that here, standard variable assignment in the metalanguage is serving the role that a \verb|typedef| declaration would fill in a C-like language.)

\textbf{Types in Ace are values in the metalanguage}, Python. More specifically, types are {\em object instances} of user-defined classes that inherit from the Ace-provided \verb|ace.Type| class. For example, \verb|global_ptr(double)| is an instance of \verb|ace.OpenCL.GlobalPtrType| instantiated with the target type, \verb|double|, as a constructor argument. The types \verb|double| and \verb|int| (imported from \verb|ace.OpenCL| on Line 4.2) are instances of \verb|ace.OpenCL.FloatType| and \verb|ace.OpenCL.IntegerType|, respectively. As we will describe further in Section \ref{att}, this notion of types as metalanguage objects is a key to Ace's flexibility.
